A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-05-18 bioc-workflow.Rmd
fixed some typos

## The use-case: prediction sub-cellular localisation in pluripotent embroyonic mouse stem cells

As a use-case we analyse a recent high-throughput spatial proteomics dataset from pluripotent mouse embryonic stem cells (E14TG2a) [@hyper]. The data was generated using hyperplexed LOPIT (hyperLOPIT), a novel technique for robust classification of protein localisation across the whole cell. The method uses an elaborate sub-cellular fractionation scheme, enabled by the use of Tandem Mass Tag (TMT) 10-plex and application of the MS data acquisition technique named synchronous precursor selection MS^3 (SPS)-MS^3 [@McAlister:2014], for TMT quantification with high accuracy and precision. Three biological replicates were generated from the E14TG2a experiment, the first was to target low density fractions and the second and third were to emphasis seperation of the denser organelles. The intersect of replicates 1 and 2 was treated as a 20-plex dataset for the analysis discussed in the manuscipt [@hyper] as it has been shown that combining replicates across from different gradients can increase spatial resolution [@trotter]. The combination of replicates resulted in 5032 proteins common in both experiments. Before combination, the two replicates were separately normalised by sum across the 10 channels (i.e. such that the sum of each protein's intensity is 1), for each replicate respectively. Normalisation is an essential part of data processing and several methods are available in `MSnbase`. The combination of data can also be performed effectively in `MSnbase` as detailed in the dedicated 'Combining MSnSet instances' section of the [`MSnbase` tutorial vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-demo.pdf).

## The infrastructre: pRoloc and MSnbase in Bioconductor

To make use of the full functionailty of the `pRoloc` software one is required to import their data into R as a `MSnSet` instance. The `MSnSet` is a dedicated data structure for the efficient manipulation and processing of mass spectrometry and proteomics data in R. Figure 1 illustrates a simplified view of the `MSnSet` structure; there exists 3 key slots (1) the `exprs` slot for storing the quantitation data, (2) the `fData` slot for storing the feature meta-data, and finally (3) the `pData` slot for storing the sample meta-data. 
![**Figure 1.** Simplified representation of the `MSnSet` data structure (reproduced with permission from the `MSnbase` vignette)](./Figures/msnset.png)

There are a number of ways to import quantitation data and create a `MSnSet` instance and all methods are described in the [`MSnbase` input/ouput capabilities vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf). One suggested simple method is to use the function `readMSnSet2` in `MSnbase`. The function takes a single spreadsheet as input and extracts the columns containing the quantitation data, as identified by the argument `ecol`, to create the expression data, while the other columns in the spreadsheet are appeneded to the feature meta-data slot. 
By example, in the code chunk below we read in the `csv` speadsheet containing the quantitation data from the intersect of replicates 1 and 2 of the mouse map [@hyper], using the `readMSnSet2` function. The data is as available online with the manuscipt (see tab 2 of the `xlsx` supplementary data set 1 in [@hyper]) and also as both a `csv` and `MSnSet` object in the Bioconductor `pRolocdata` data package.

To use the `readMSnSet2` function, as a minimum one must specify the file path to the data and which columns of the spreadsheet contain quantitation data. The `getEcols` function exists to help users identify which columns of the spreadsheet that contain the quantitation data. The spreadsheet of E14TG2a data; "hyperLOPIT-SIData-ms3-rep12-intersect.csv", contains 5032 proteins common across the 2 biological replicates for the respective 2 x 10-plex reporter tags, along with assoiated feature meta-data such as protein markers, protein description, number of quantified peptides etc. The spreadsheet contains two headers, with the second header containing information about where the quantitation data is stored. We can display the names of the second header by calling the `getEcols` function with the argument `n = 2`, to specify that we wish to display the column names of the second header. It is now easy for one to identify that the quantitation data is located in columns 8 to 27. It is also possible to pass the optional argument `fnames` to indicate which column to use as the labels by which to identify each protein in the sample. Here, we use `fnames = 1` to use the Uniprot identifiers contained in the first column of the speadsheet.

@@

As briefly mentioned above, the quantitation data is stored in the `exprs` slot of the `MSnSet` and can be accessed by `exprs(lopit2015)`. The feature meta-data is stored in the `fData` slot and can be accessed by `fData(lopit2015)`. When using `readMSnSet2`, automatically, everything that is not defined as quantitation data by `ecol` or the fetaure names by `fnames` is deposited to the `fData` slot. As we wish to demonstrate the complete analysis of this data we remove the results from the prior analysis described in [@hyper] in the code chunk below. We see the `fData` contains 25 columns describing information such as the number of peptides, associated markers, machine learning results etc. For demonstration in the code chunk below keep the 2nd, 8th and 11th columns which contain the Uniprot entry names and two different marker sets to use an input for machine learning analyses (see sections on markers and subsequent sections).

@@

Data quality is routinely examined through visualisation to verify that sub-cellular niches have been separated along the gradient. Based on De Duve's principle [@DeDuve:1981] proteins that co-localise exhibit similar quantitation profiles across the gradient fractions employed. One approach that has been widely used to visualise and inspect highthrouput mass spectrometry-based proteomics data is principal components analysis (PCA). PCA is one of many dimensionality reduction methods, that allow one to effectively summarise multi-dimensional data in to 2 or 3 dimensions to enable visualisation. Very generally, the original continuous multi-dimensional data is transformed into a set of orthogonal components ordered according to the amount of variability that they describe. The `plot2D` method in `pRoloc` allows one to plot the principal components (PCs) of a dataset against one another, by default the first two components are plotted on the x- and y-axis, respectively (the `dims` argument can be used to plot other PCs). If distinct clusters are observed, we assume that there is organellar separation present in the data. In the code chunk below we produce a PCA plot of the mouse stem cell dataset. One point on the plot represents one protein. We can indeed see several distinct protein clusters. Although, plotting the PCs does not give us a hard quantitative measure of separation, it is extremely useful summarising complex experimental information in one figure, to get an simplified overview of the data.

@@

In the context of spatial proteomics, a marker protein is defined as a well-known resident of a specific sub-cellular niche in a species *and* condition of interest. Applying this to machine learning (ML), and specifically supervised learning, for the task of protein localisation prediction, markers constitute the labelled training data to use as input for a classification analyses. Defining well-known residents, and obtaining labelled training data for ML analyses can be time consuming, but is important to define markers that are representative of the multivariate data space and on which a classifier will be trained and generated. `pRoloc` provides a convenience function, `addMarkers`, to directly add markers to a `MSnSet` object, as demonstrated in the code chunk below. These marker sets can be accessed using the `pRolocmarkers()` function. Marker sets are stored as a simple named vector in R, and originate from inhouse user-defined spreadsheets or a set of markers from previous published studies. The marker vectors that can be accessed from `pRolocmarkers` are named vectors and to enable mapping between the markers and the `MSnSet` instance it is required that the `featureNames` of the `MSnSet` instance match the `names` of the marker. The mouse dataset used here has Uniprot IDs stored as the `featureNames` (see `head(featureNames(lopit2015))`) and the names of the vector of the mouse markers (`mmus` markers) are Uniprot entry names (see `head(mrk)` in the code chunk below), therefore it is required we update the `MSnSet` to match the names of the markers, as demonstrated below. It is then possible to match names between the markers and the `MSnSet` instance. We see below that the markers cover many sub-cellular niches, with many niches only containing a few proteins. Depending on the biological question and downstream analyses we may wish the subset these marker classes, this can be done using the `minMarkers` function. In the code chunk below, we demonstrate how to add markers using `pRolocmarkers` function and then visualise these annotations using the `plot2D` function.

@@

The extraction of sub-cellular protein clusters can be difficult owing to the limited number of marker proteins that exist in databases and elsewhere. Furthermore, given the vast complexity of the cell, automatic annotation retrieval does not always give a full representation of the true sub-cellular diversity in the data. For downstream analyses, such as supervised machine learning, it is desirable to obtain reliable markers that cover as many sub-cellular niches as possible, as these markers are directly used in the training phase of the ML classification. We find that a lack of sub-cellular diveristy in the labelled training data leads to prediction errors, as unlabelled instance can only to predicted to a class that exists in the training data  [@Breckels:2013]. In such scenarios novelty detection can be useful to identify data-specific sub-cellular groupings such as organelles and protein complexes. The phenotype discovery (phenoDisco) algorithm [@Breckels:2013] is one such method and available in `pRoloc`. It is an iterative semi-supervised learning method that combines the classification of proteins on existing labelled data with the detection of new clusters. Novelty detection methods are also useful for confirming the presence of known clusters in an unbiased fashion. For example, in [@hyper] the `phenoDisco` algorithm was used to detect the data-specific confirmation and presence of the nucleus and sub-compartments, of it was expected that nucleus associated clusters existed within the data as the protocol included a separate chromatin-rich fraction to enrich for nuclei. In the code chunk below, we show how to run a a series of novelty detection experiments using the `phenoDisco` algorithm on the mouse stem-cell dataset. There are several optional arguments that can be set dependent on the type of biological question of interest.

>>>>

## The use-case: prediction sub-cellular localisation in pluripotent embryonic mouse stem cells

As a use-case we analyse a recent high-throughput spatial proteomics dataset from pluripotent mouse embryonic stem cells (E14TG2a) [@hyper]. The data was generated using hyperplexed LOPIT (hyperLOPIT), a novel technique for robust classification of protein localisation across the whole cell. The method uses an elaborate sub-cellular fractionation scheme, enabled by the use of Tandem Mass Tag (TMT) 10-plex and application of the MS data acquisition technique named synchronous precursor selection MS^3 (SPS)-MS^3 [@McAlister:2014], for TMT quantification with high accuracy and precision. Three biological replicates were generated from the E14TG2a experiment, the first was to target low density fractions and the second and third were to emphasis separation of the denser organelles. The intersect of replicates 1 and 2 was treated as a 20-plex dataset for the analysis discussed in the manuscript [@hyper] as it has been shown that combining replicates across from different gradients can increase spatial resolution [@trotter]. The combination of replicates resulted in 5032 proteins common in both experiments. Before combination, the two replicates were separately normalised by sum across the 10 channels (i.e. such that the sum of each protein's intensity is 1), for each replicate respectively. Normalisation is an essential part of data processing and several methods are available in `MSnbase`. The combination of data can also be performed effectively in `MSnbase` as detailed in the dedicated 'Combining MSnSet instances' section of the [`MSnbase` tutorial vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-demo.pdf).

## The infrastructure: pRoloc and MSnbase in Bioconductor

To make use of the full functionality of the `pRoloc` software one is required to import their data into R as a `MSnSet` instance. The `MSnSet` is a dedicated data structure for the efficient manipulation and processing of mass spectrometry and proteomics data in R. Figure 1 illustrates a simplified view of the `MSnSet` structure; there exists 3 key slots (1) the `exprs` slot for storing the quantitation data, (2) the `fData` slot for storing the feature meta-data, and finally (3) the `pData` slot for storing the sample meta-data. 
![**Figure 1.** Simplified representation of the `MSnSet` data structure (reproduced with permission from the `MSnbase` vignette)](./Figures/msnset.png)

There are a number of ways to import quantitation data and create a `MSnSet` instance and all methods are described in the [`MSnbase` input/output capabilities vignette](http://bioconductor.org/packages/release/bioc/vignettes/MSnbase/inst/doc/MSnbase-io.pdf). One suggested simple method is to use the function `readMSnSet2` in `MSnbase`. The function takes a single spreadsheet as input and extracts the columns containing the quantitation data, as identified by the argument `ecol`, to create the expression data, while the other columns in the spreadsheet are appended to the feature meta-data slot. 
By example, in the code chunk below we read in the `csv` spreadsheet containing the quantitation data from the intersect of replicates 1 and 2 of the mouse map [@hyper], using the `readMSnSet2` function. The data is as available online with the manuscript (see tab 2 of the `xlsx` supplementary data set 1 in [@hyper]) and also as both a `csv` and `MSnSet` object in the Bioconductor `pRolocdata` data package.

To use the `readMSnSet2` function, as a minimum one must specify the file path to the data and which columns of the spreadsheet contain quantitation data. The `getEcols` function exists to help users identify which columns of the spreadsheet that contain the quantitation data. The spreadsheet of E14TG2a data; "hyperLOPIT-SIData-ms3-rep12-intersect.csv", contains 5032 proteins common across the 2 biological replicates for the respective 2 x 10-plex reporter tags, along with associated feature meta-data such as protein markers, protein description, number of quantified peptides etc. The spreadsheet contains two headers, with the second header containing information about where the quantitation data is stored. We can display the names of the second header by calling the `getEcols` function with the argument `n = 2`, to specify that we wish to display the column names of the second header. It is now easy for one to identify that the quantitation data is located in columns 8 to 27. It is also possible to pass the optional argument `fnames` to indicate which column to use as the labels by which to identify each protein in the sample. Here, we use `fnames = 1` to use the Uniprot identifiers contained in the first column of the spreadsheet.

@@

As briefly mentioned above, the quantitation data is stored in the `exprs` slot of the `MSnSet` and can be accessed by `exprs(lopit2015)`. The feature meta-data is stored in the `fData` slot and can be accessed by `fData(lopit2015)`. When using `readMSnSet2`, automatically, everything that is not defined as quantitation data by `ecol` or the feature names by `fnames` is deposited to the `fData` slot. As we wish to demonstrate the complete analysis of this data we remove the results from the prior analysis described in [@hyper] in the code chunk below. We see the `fData` contains 25 columns describing information such as the number of peptides, associated markers, machine learning results etc. For demonstration in the code chunk below keep the 2nd, 8th and 11th columns which contain the Uniprot entry names and two different marker sets to use an input for machine learning analyses (see sections on markers and subsequent sections).

@@

Data quality is routinely examined through visualisation to verify that sub-cellular niches have been separated along the gradient. Based on De Duve's principle [@DeDuve:1981] proteins that co-localise exhibit similar quantitation profiles across the gradient fractions employed. One approach that has been widely used to visualise and inspect high throughput mass spectrometry-based proteomics data is principal components analysis (PCA). PCA is one of many dimensionality reduction methods, that allow one to effectively summarise multi-dimensional data in to 2 or 3 dimensions to enable visualisation. Very generally, the original continuous multi-dimensional data is transformed into a set of orthogonal components ordered according to the amount of variability that they describe. The `plot2D` method in `pRoloc` allows one to plot the principal components (PCs) of a dataset against one another, by default the first two components are plotted on the x- and y-axis, respectively (the `dims` argument can be used to plot other PCs). If distinct clusters are observed, we assume that there is organellar separation present in the data. In the code chunk below we produce a PCA plot of the mouse stem cell dataset. One point on the plot represents one protein. We can indeed see several distinct protein clusters. Although, plotting the PCs does not give us a hard quantitative measure of separation, it is extremely useful summarising complex experimental information in one figure, to get an simplified overview of the data.

@@

In the context of spatial proteomics, a marker protein is defined as a well-known resident of a specific sub-cellular niche in a species *and* condition of interest. Applying this to machine learning (ML), and specifically supervised learning, for the task of protein localisation prediction, markers constitute the labelled training data to use as input for a classification analyses. Defining well-known residents, and obtaining labelled training data for ML analyses can be time consuming, but is important to define markers that are representative of the multivariate data space and on which a classifier will be trained and generated. `pRoloc` provides a convenience function, `addMarkers`, to directly add markers to a `MSnSet` object, as demonstrated in the code chunk below. These marker sets can be accessed using the `pRolocmarkers()` function. Marker sets are stored as a simple named vector in R, and originate from in-house user-defined spreadsheets or a set of markers from previous published studies. The marker vectors that can be accessed from `pRolocmarkers` are named vectors and to enable mapping between the markers and the `MSnSet` instance it is required that the `featureNames` of the `MSnSet` instance match the `names` of the marker. The mouse dataset used here has Uniprot IDs stored as the `featureNames` (see `head(featureNames(lopit2015))`) and the names of the vector of the mouse markers (`mmus` markers) are Uniprot entry names (see `head(mrk)` in the code chunk below), therefore it is required we update the `MSnSet` to match the names of the markers, as demonstrated below. It is then possible to match names between the markers and the `MSnSet` instance. We see below that the markers cover many sub-cellular niches, with many niches only containing a few proteins. Depending on the biological question and downstream analyses we may wish the subset these marker classes, this can be done using the `minMarkers` function. In the code chunk below, we demonstrate how to add markers using `pRolocmarkers` function and then visualise these annotations using the `plot2D` function.

@@

The extraction of sub-cellular protein clusters can be difficult owing to the limited number of marker proteins that exist in databases and elsewhere. Furthermore, given the vast complexity of the cell, automatic annotation retrieval does not always give a full representation of the true sub-cellular diversity in the data. For downstream analyses, such as supervised machine learning, it is desirable to obtain reliable markers that cover as many sub-cellular niches as possible, as these markers are directly used in the training phase of the ML classification. We find that a lack of sub-cellular diversity in the labelled training data leads to prediction errors, as unlabelled instance can only to predicted to a class that exists in the training data  [@Breckels:2013]. In such scenarios novelty detection can be useful to identify data-specific sub-cellular groupings such as organelles and protein complexes. The phenotype discovery (phenoDisco) algorithm [@Breckels:2013] is one such method and available in `pRoloc`. It is an iterative semi-supervised learning method that combines the classification of proteins on existing labelled data with the detection of new clusters. Novelty detection methods are also useful for confirming the presence of known clusters in an unbiased fashion. For example, in [@hyper] the `phenoDisco` algorithm was used to detect the data-specific confirmation and presence of the nucleus and sub-compartments, of it was expected that nucleus associated clusters existed within the data as the protocol included a separate chromatin-rich fraction to enrich for nuclei. In the code chunk below, we show how to run a a series of novelty detection experiments using the `phenoDisco` algorithm on the mouse stem-cell dataset. There are several optional arguments that can be set dependent on the type of biological question of interest.
